### 2013-06-05 text/Chapter02.tex
Finished pathways, copy & past chapters 3 and 4.

We demonstrate the problem by looking at input validation. In distributed applications 
that communicate over unsecured HTTP channel, user input needs to be validated interactively
on the client-side (to provide immediate response) and then again on the server-side (to 
guarantee safety). For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\
\quad\ident{name} \neq \str{""} ~~\&\&~~ \ident{forall~isLetter~name}
\\[0.5em]
\kvd{let}~\ident{displayProduct} = \lambda \ident{name} \rightarrow\\
\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{displayProductPage~name}\\
\quad\kvd{else}~\ident{displayErrorPage}~() 
\end{array}
\end{equation*}
%
The function \ident{validateInput} can be compiled to both JavaScript (for client-side) and
native code (for server-side). However, \ident{displayProduct} uses other functionality
(generating web pages) that is only available on the server-side, so it can only be compiled to
native code.

@@

@@

% \paragraph{Safe locking}
% Flanagan and Abadi \cite{app-safe-locking}
%
% => This has coeffect style judgments, but it has effect-style lambda
%

@@

\paragraph{Software updating}
Dynamic software updating (DSU) is the ability to update programs at runtime without stopping them.

Stoyle et al. \cite{app-dsu-mutatis}.

\paragraph{Thesis perspective}
?

\newpage
%===================================================================================================

@@

\section{Summary}
Oops!

>>>>

We demonstrate the problem by looking at input validation. In distributed applications 
that communicate over unsecured HTTP channel, user input needs to be validated interactively
on the client-side (to provide immediate response) and then again on the server-side (to 
guarantee safety). For example:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{validateInput} = \lambda \ident{name} \rightarrow\\
\quad\ident{name} \neq \str{""} ~~\&\&~~ \ident{forall~isLetter~name}
\\[0.5em]
\kvd{let}~\ident{displayProduct} = \lambda \ident{name} \rightarrow\\
\quad\kvd{if}~\ident{validateInput name}~\kvd{then}~\ident{displayProductPage~name}\\
\quad\kvd{else}~\ident{displayErrorPage}~() 
\end{array}
\end{equation*}
%
The function \ident{validateInput} can be compiled to both JavaScript (for client-side) and
native code (for server-side). However, \ident{displayProduct} uses other functionality
(generating web pages) that is only available on the server-side, so it can only be compiled to
native code.

@@

\paragraph{Safe locking}
In the previous examples, the context provides additional values or functions that may be accessed
at runtime. However, it may also track \emph{permissions} to perform some operation. This is done
in the type system for safe locking of Flanagan and Abadi \cite{app-safe-locking}.

The system prevents race conditions (by only allowing access to mutable state under a lock)
and avoids deadlocks (by imposing strict partial order on locks). The following 
program uses a mutable state under a lock:
%
\begin{equation*}
\begin{array}{l}
\kvd{newlock}~l:\rho~\kvd{in}\\
\kvd{let}~\ident{state}~=~\ident{ref}_\rho~10~\kvd{in}\\
\kvd{sync}~l~(!\ident{state})
\end{array}
\end{equation*}
%
The declaration \kvd{newlock} creates a lock $l$ protecting memory region $\rho$. We can than
allocate mutable variables in that memory region (second line). An access to mutable variable
is only allowed in scope that is protected by a lock. This is done using the \kvd{sync} keyword,
which locks a lock and evaluates an expression in a context that contains permission to access
memory region of the lock ($\rho$ in the above example).

The type system for safe locking associates the list of permission with the variable context.
It uses judgements of a form $\Gamma, m \vdash e : \alpha$ specifying that an expression has a
type in context $\Gamma$, given permissions (a list of locked regions) $m$. However, the treatment
of lambda abstraction differs from the one for implicit parameters or rebindable resources.
In the system for locking, code inside lambda function cannot use permissions from the scope
where the function is declared. This is a necessary requirement -- a lambda function created 
under a lock cannot access protected memory, because it will be executed later. We discuss how
this restriction fits into our general coeffect framework in Section~X.Y.

%
% => This has coeffect style judgments, but it has effect-style lambda
%

@@

\paragraph{Software updating}
Dynamic software updating (DSU) \cite{app-dsu-programs,app-dsu} is the ability to update programs at
runtime without stopping them. The Proteus system developed by Stoyle et al. \cite{app-dsu-mutatis} 
investigates what language support is needed to enable safe dynamic software updating in C-like 
languages. The system is based on the idea of capabilities.

The system distinguishes between \emph{concrete} uses and \emph{abstract} uses of a value. When
a value is used concretely, the program examines its representation (and so it is not safe to
change the representation during an update). An abstract use of a value does not need to examine
the representation and so updating the value does not break the program.

The Proteus system uses capabilities to restrict what types may be used concretely after any point
in the program. All other types, not listed in the capability, can be dynamically updated as this
will not change concrete representation of types accessed later in the evaluation.

Similarly to Capability Calculus, capabilities in DSU can be changed by a function call. For 
example, calling a function that may update certain types makes it impossible to use those types
concretely following the function call. This means that DSU uses the context \emph{actively}
and not just \emph{passively}.

@@

As demonstrated in this section, there is a huge number of systems and applications that exhibit
a form of context-dependence. The range includes different static analyses (liveness, provenance), 
well-known programming language features (implicit parameters and type classes) as well as features
not widely available (e.g. for distributed programming).

It is impossible to cover all of these topics in a single coherent thesis and so we focus on 
two key aspects:

\begin{compactitem}
\item \textbf{Flat vs. structural.} We look at both flat coeffects (single value for entire context) and 
  structural coeffects (single value per variable). We use liveness, implicit parameters and 
  data-flow to introduce flat coeffects (Section~X) and liveness, refined data-flow and tainting
  to talk about structural coeffects (Section~Y).

\item \textbf{Analysis vs. restriction.} Some of the discussed examples can be viewed as static
  analyses that obtain some information about programs (i.e.~the number of required past values
  in data-flow). Other examples provide type system that rules out certain invalid programs 
  (e.g.~safe locking). We cover this topic when discussing \emph{partial coeffects} in Section~Z.

\item \textbf{May vs. must analysis.} When discussing liveness, we observed that we can obtain 
  two different analyses depending on how conditionals are treated. We discuss this topic in 
  Section~X. % TBD - could be a chapter
\end{compactitem}

Although we also looked at examples of \emph{active} contextual computations (where developers can
write functions that modify the context), we do not consider these applications, to keep the 
material presented in this thesis focused. We briefly discuss them as future work in Section~X.

@@

\section{Summary}

This chapter presented four different pathways leading to the idea of coeffects. We also 
introduced the most important related work, although presenting related work was not the
main goal of the chapter. The main goal was to show the idea of coeffects as a logical follow up
to a number of research directions. For this reason, we highlighted only certain aspects of 
related work -- the remaining aspects as well as important technical details are covered 
in later chapters.

The first pathway looks at applications and systems that involve notion of \emph{context}.
The two coeffect calculi we present aim to unify some of these systems. The second pathway
follows as a dualization of well-known effect systems. However, this is not simply a syntactic
transformation, because coeffect systems treat lambda abstraction differently. The third 
pathway follows by extending comonadic semantics of context-dependent computations with 
indexing and building a type system analogous to effect system from the ``marriage of effects
and monads''. Finally, the fourth pathway starts with sub-structural type systems. Coeffect
systems naturally arise by annotating bunches in bunched logics with additional information.
